Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-03-25 Paper/manuscript_v2/manuscript_07_01.tex
Minor edits and added to dos to Section 2

In this work, GPs will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \bm{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n|f(\bm{x}_n),\phi)$, where $p$ is some parametric distribution with parameters $f$ and $\phi$, and $f$ is an unknown function with GP prior, which depends on an input $\bm{x}_n\in {\rm I\!R}^D$. This generalizes readily to more complex models depending on several unknown functions, for example such as $p(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or multilevel models. Our goal is to obtain posterior distribution for the value of the function $\tilde{f}=f(\tilde{\bm{x}})$  evaluated at a new input point $\tilde{\bm{x}}$.
% That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.\\

@@

If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),\sigma)$ then $\bm{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GPs and $O(nm^2)$ for sparse GPs). If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or $p(y_n|f(\bm{x}_n),\phi)$ is non-Gaussian, the marginalization does not have closed form solution. Furthermore, if a prior distribution is imposed on $\phi$ and $\bm{\theta}$ to form a joint posterior for $\phi$, $\bm{\theta}$ and $\bm{f}$, approximate inference such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011}, Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient} need to be used. In this paper, we focus on the use of MCMC for integrating over the joint
posterior. MCMC is not usually the fastest approach, but allows
accurate inference and uncertainty estimates for general models in probabilistic programming
settings. We consider the computational costs of GPs specifically from
this point of view.

\subsection{Covariance function and spectral density}\label{ch4_sec_cov}

The covariance function is the crucial ingredient in a GP as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A covariance function needs be symmetric and positive semi-definite \citep{rasmussen2006gaussian}. A stationary covariance function is a function of $\bm{\tau}=\bm{x}-\bm{x}' \in {\rm I\!R}^D$, such that it can be written $k(\bm{x},\bm{x}') = k(\bm{\tau})$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\bm{x},\bm{x}') = k(|\bm{x}-\bm{x}'|) = k(r), r\in {\rm I\!R}$, which means that the covariance is both translation and rotation invariant. The most commonly used distance between observations is the norm L2 $(|\bm{x}-\bm{x}'|_{L2})$, also known as Euclidean distance, although other types of distances can be considered.

@@

\noindent where $\nu$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp\!\! \big(\! -\frac{1}{2} \frac{r^2}{\ell^2}\big),  \\
k_{\frac{3}{2}}(r)&=\alpha\Big(1+\frac{\sqrt{3}r}{\ell}\Big) \! \exp\!\! \big(\! -\frac{\sqrt{3}r}{\ell}\big). 
\end{align*}

\noindent The former is commonly known as squared exponential (exponentiated quadratic) covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form:
%
\begin{align*}
k_{\infty}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \exp\!\!\Big(\! -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\Big),\\
k_{\frac{3}{2}}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \bigg(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\bigg) \! \exp\!\!\Big(\! - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\Big). \\
\end{align*}

\noindent Notice that the previous expressions have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. The use of a multidimensional length-scale basically turns the isotropic covariance function into a non-isotropic one. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

@@

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ is a distance in the frequency domain, and $\ell$ and $\alpha$ are the lengthscale and magnitude (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form:
%
\begin{align}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^D  \ell^D \exp(-0.5 \ell^2 \omega^2), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\,\ell^3}\Big(\frac{3}{\ell^2}+\omega^2 \Big)^{\!-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}

\noindent Particularizing to an input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form:
%
\begin{align*}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \exp\!\!\big(\!-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \big),   \\
S_{\frac{3}{2}}(\omega)&= \alpha \, 32\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\big(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \big)^{\!-3}.
\end{align*}

>>>>

In this work, GPs will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \bm{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n|f(\bm{x}_n),\phi)$, where $p$ is some parametric distribution with parameters  $\phi$, and $f$ is an unknown function with GP prior, which depends on an input $\bm{x}_n\in {\rm I\!R}^D$. This generalizes readily to more complex models depending on several unknown functions, for example such as $p(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or multilevel models. Our goal is to obtain the posterior distribution for the value of the function $\tilde{f}=f(\tilde{\bm{x}})$  evaluated at a new input point $\tilde{\bm{x}}$.
% That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.\\

@@

If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),\sigma)$ then $\bm{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GPs and $O(nm^2)$ for sparse GPs). If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),g(\bm{x}_n))$ \todo{Is something missing here? The function g is missing on the left hand side. } or $p(y_n|f(\bm{x}_n),\phi)$ is non-Gaussian, the marginalization does not have a closed form solution. Furthermore, if a prior distribution is imposed on $\phi$ and $\bm{\theta}$ to form a joint posterior for $\phi$, $\bm{\theta}$ and $\bm{f}$, approximate inference such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011}, Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient} are required. In this paper, we focus on the use of MCMC for integrating over the joint
posterior. MCMC is usually not the fastest approach, but it is flexible and allows
accurate inference and uncertainty estimates for general models in probabilistic programming
settings. We consider the computational costs of GPs specifically from
this point of view.

\subsection{Covariance functions and spectral density}\label{ch4_sec_cov}

The covariance function is the crucial ingredient in a GP as it encodes our prior assumptions about the function, and defines a correlation structure which characterizes the correlations between function values at different inputs. A covariance function needs be symmetric and positive semi-definite \citep{rasmussen2006gaussian}. A stationary covariance function is a function of $\bm{\tau}=\bm{x}-\bm{x}' \in {\rm I\!R}^D$, such that it can be written $k(\bm{x},\bm{x}') = k(\bm{\tau})$, which means that the covariance is invariant to translations. Isotropic covariance functions are only depends on the input points through the norm of the difference, $k(\bm{x},\bm{x}') = k(|\bm{x}-\bm{x}'|) = k(r), r\in {\rm I\!R}$, which means that the covariance is both translation and rotation invariant. The most commonly used distance between observations is the L2-norm $(|\bm{x}-\bm{x}'|_{L2})$, also known as Euclidean distance, although other types of distances can be considered.

@@

\noindent where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp\!\! \big(\! -\frac{1}{2} \frac{r^2}{\ell^2}\big),  \\
k_{\frac{3}{2}}(r)&=\alpha\Big(1+\frac{\sqrt{3}r}{\ell}\Big) \! \exp\!\! \big(\! -\frac{\sqrt{3}r}{\ell}\big). 
\end{align*}

\noindent The former is commonly known as the squared exponential (exponentiated quadratic) covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form
%
\begin{align*}
k_{\infty}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \exp\!\!\Big(\! -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\Big),\\
k_{\frac{3}{2}}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \bigg(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\bigg) \! \exp\!\!\Big(\! - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\Big). \\
\end{align*}

\noindent Notice that the previous expressions have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns the isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

@@

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ denotes the frequency, and $\ell$ and $\alpha$ are the lengthscale and magnitude (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form
%
\begin{align}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^D  \ell^D \exp(-0.5 \ell^2 \omega^2), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\,\ell^3}\Big(\frac{3}{\ell^2}+\omega^2 \Big)^{\!-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}
\todo{Is it only purpose that only some equations have numbers?}
\noindent For input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \exp\!\!\big(\!-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \big),   \\
S_{\frac{3}{2}}(\omega)&= \alpha \, 32\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\big(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \big)^{\!-3}.
\end{align*}
